A pandemic-scale Ancestral Recombination Graph for SARS-CoV-2

### 2025-09-25 latex/main.tex
Updates from Overleaf
Pass on "Overview of sc2ts"

Millions of SARS-CoV-2 genome sequences were collected in
response to the COVID-19 pandemic, forming a dataset of unprecedented
depth and richness.
Estimated genealogies are fundamental to
gleaning information from this ocean of data, 
and form the primary input to many downstream analyses.
A basic assumption of methods to infer genealogies from viral 
genetic sequence data is that recombination is negligible and 
the genealogy is a tree.
Recombinant lineages have risen to global prevalence in SARS-CoV-2,
and therefore such single-tree representations are necessarily
incomplete and potentially misleading.
We present sc2ts, a method to infer reticulate genealogies 
as an Ancestral Recombination Graph (ARG) for SARS-CoV-2
that automatically detects and incorporates recombination.  
We use sc2ts to infer an ARG of 2.48 million genomes,
and show that it captures a broad range of 
known evolutionary features and provides novel 
insights into the relationships between previously identified
recombinant lineages.
Together, the inference method, infererred ARG and 
suite of open-source analysis tools, 
represent a powerful resource for future research.

@@

Here we present sc2ts, a new method to infer ARGs for SARS-CoV-2  
that incrementally adds samples over time in a similar manner 
to UShER, but also automatically detects and incorporates recombination
in real-time. Based on the Viridian dataset, a freely available  
resource purged of various systematic errors and artifacts \cite{Hunt2024},
we infer an ARG with 2.48 million whole genome sequences.
We show that this ARG is highly congruent with the UShER phylogeny built 
from the same dataset as well as the Pango lineage naming system\cite{Rambaut2020},
and also accurately captures evolutionary features such as mutational 
spectra\cite{Bloom2023} and deletions\cite{Li2023indels}.
In addition, we show that sc2ts automatically and accurately detects details about
known recombinant lineages that were painstakingly collated in a community 
process, and provides novel insights into the relationships between these
recombinant lineages. By fully incorporating these and other novel
recombination events, the resulting data structure---representing
phylogenetic relationships, point mutations, deletions, and 
recombination---provides a basis for the study of SARS-CoV-2
that systematically accounts for all of these evolutionary forces.
This ARG, coupled with the rich suite of supporting software
in the tskit\cite{Ralph2020} and VCF-Zarr~\cite{Czech2025} ecosystems,
enables pandemic-scale analysis on a laptop. 
% Dunno about this! Feels important to finish on a wider note, helping
% the reader understand the significance of all of this?
These resources
will help us to learn from the vast troves of data collected during 
pandemic, and to prepare for the next.

@@

\subsection*{Brief description of sc2ts}

Sc2ts reconstructs an ARG by incrementally adding batches of sequences as they become available
(Figure~\ref{fig:methodology}; for details, see STAR Methods).
First, sc2ts infers likely ``copying paths'' connecting each sample in the batch to node(s) in the current ARG
(allowing recombination) using the Li and Stephens (LS) model \cite{LiStephens2003},
which is a Hidden Markov Model (HMM) widely used in human genomics
to approximate the effects of mutation and recombination on genetic inheritance.
As recombination is relatively rare in viruses,
we would expect most samples to copy from a single parent in the ARG.
Recombinant samples, however, copy from two or more parents, with switching (recombination) in the copying path
between the parents resulting in different genomic segments inherited from the different parents.
Next, for clusters within the batch that are inferred to copy from the same ARG node,
sc2ts uses standard phylogenetic techniques to resolve relationships within the cluster.
Finally, sc2ts attaches the phylogenies of the clusters to the current ARG, and
applies parsimony-improving heuristics to address issues introduced by the inherent greediness of this strategy.
[TODO add brief summary of sample filtering and how we handle VOCs, 
refing Methods for details]

@@

%% JK: This is copy of old Methods section "overview of sc2ts" which is basically
%% the same as this section. TODO Merge the useful text in here.

% Sc2ts (pronounced ``scoots'', optionally)
% is a method for inferring Ancestral Recombination Graphs (ARGs)
% from densely sampled genome sequence data in real time,
% in which recombination occurs at a low but significant rate.
% The basic idea is to incrementally update an ARG regularly
% with the sequences collected since the previous batch (Figure
% \ref{fig:methodology}).
% We used daily batches for SARS-CoV-2,
% and use days as the time interval for clarity in what follows.
% The first step is to find likely ``copying paths'' under the Li and Stephens (LS) model \cite{LiStephens2003}
% for each sequence in the daily batch to the current ARG.
% These copying paths will mostly consist of a new sample sequence copying
% from a node in the ARG with a small number of mutations,
% and often many samples from a daily batch will copy from the same ARG node.
% The second step is then to ``resolve'' these implied polytomies
% by using a standard neighbor-joining algorithm \cite{Saitou1987}.
% This greedy update strategy inevitably leads to unparsimonious topologies,
% and the third step is to then increase the overall parsimony of the inferred ARG
% by making some simple topological updates.
% Recombination is inferred as an integral and ongoing part of this process,
% requiring only a few additional steps to facilitate later analysis.
% The result of inference for a given day is then a genealogy recording genetic inheritance
% as well as mutation and recombination events
% for all of the samples added to the ARG up to that day.

% Sc2ts is implemented in Python building on core PyData infrastructure
% such as Numpy\cite{Harris2020array}, SciPy\cite{Virtanen2020scipy}
% and Pandas~\cite{Mckinney2010data} ,
% using pyfaidx\cite{Shirley2015efficient} to access alignment data
% and Numba~\cite{Lam2015numba} to accelerate certain alignment processing
% tasks.
% [TODO JK to fill this out]

@@

Furthermore, some of the Pango X lineages above originate from
shared recombination events in the ARG (Table~\ref{tab:pango_x_lineages}).
An example of a set of ``nested'' closely related lineages
consists of XZ, XAC, XAD, XAE, and XAP,
which descend from a single recombination event rather than five independent
events in the ARG (Figure~\ref{fig:subgraphs}B), and
which we tentatively group using the unused label Xx.
These Pango X lineages were designated around the same time.
These results highlight that the ARG can help
clarify the relationships between recombinants,
allowing more precise interpretation of evidence for recombination.

>>>>

Millions of SARS-CoV-2 genome sequences were collected in
response to the COVID-19 pandemic, forming a dataset of unprecedented
depth and richness.
Estimated genealogies are fundamental to
gleaning information from this ocean of data, 
and form the primary input to many downstream analyses.
A basic assumption of methods to infer genealogies from viral 
genetic sequence data is that recombination is negligible and 
the genealogy is a tree.
Recombinant lineages have risen to global prevalence in SARS-CoV-2,
and therefore such single-tree representations are necessarily
incomplete and potentially misleading.
We present sc2ts, a method to infer reticulate genealogies 
as an Ancestral Recombination Graph (ARG) for SARS-CoV-2
that automatically detects and incorporates recombination.  
We use sc2ts to infer an ARG of 2.48 million genomes,
and show that it captures a broad range of 
known evolutionary features and provides novel 
insights into the relationships between previously identified
recombinant lineages.
Together, the inference method, inferred ARG and 
suite of open-source analysis tools, 
represent a powerful resource for future research.

@@

Here we present sc2ts, a new method to infer ARGs for SARS-CoV-2  
that incrementally adds samples over time in a similar manner 
to UShER, but also automatically detects and incorporates recombination
in real-time. Based on the Viridian dataset, a freely available  
resource purged of various systematic errors and artifacts \cite{Hunt2024},
we infer an ARG with 2.48 million whole genome sequences.
We show that this ARG is highly congruent with the UShER phylogeny built 
from the same dataset as well as the Pango lineage naming system\cite{Rambaut2020},
and also accurately captures evolutionary features such as mutational 
spectra\cite{Bloom2023} and deletions\cite{Li2023indels}.
In addition, we show that sc2ts automatically and accurately detects details about
known recombinant lineages that were painstakingly collated in a community 
process, and provides novel insights into the relationships between these
recombinant lineages. By fully incorporating these and other novel
recombination events, the resulting data structure---representing
phylogenetic relationships, point mutations, deletions, and 
recombination---provides a basis for the study of SARS-CoV-2
that systematically accounts for all of these evolutionary forces.
This ARG, coupled with the rich suite of supporting software
in the tskit\cite{Ralph2020} and VCF-Zarr~\cite{Czech2025} ecosystems,
enables pandemic-scale analysis on a laptop. 
% Dunno about this! Feels important to finish on a wider note, helping
% the reader understand the significance of all of this?
These resources
will help us to learn from the vast troves of data collected during the
pandemic, and to prepare for the next.

@@

\subsection*{Overview of sc2ts}

@@

Sc2ts reconstructs an ARG by incrementally adding daily batches of 
sequences (Figure~\ref{fig:methodology}).
First, sc2ts infers likely ``copying paths'' connecting each sample in the 
batch to node(s) in the current ARG
(allowing recombination) using a simplified version of the Li and Stephens (LS) model
\cite{LiStephens2003}. 
% Important to go into a bit of detail about LS here as it's so fundamental
% and novel in the area.
The LS model is a Hidden Markov Model (HMM) widely used in human genomics
to approximate the effects of mutation and recombination on genetic
inheritance\citep{Delaneau2019-wl,Browning2018-nk},
typically parameterised by population-scaled mutation and
recombination rates varying along the genome~\cite{Donnelly2010-coalescent}.
In the interest of interpretability, we simplified this model to 
use a single parameter $k$ which essentially 
controls the number of recurrent mutations that is preferred to 
switching to a different parent via recombination. We used the 
highly efficient LS HMM implementation in tsinfer\cite{Kelleher2019},
with some additional optimisations enabled by our simplified model,
to find exact Viterbi solutions of these copying paths for each sample
given the current ARG.
Note that this is a form of parsimony, with the recombination
penalty $k$ determining the relative parsimoniousness of recombinant
vs non-recombinant solutions to the HMM.
The per-sample copying paths produced by the HMM (most of which do not 
involve recombination) form clusters of samples within a batch, which 
are then resolved using standard phylogenetic techniques.
Finally, sc2ts attaches the phylogenies of the clusters to the current ARG, and
applies parsimony-improving heuristics to address issues introduced 
by the inherent greediness of this strategy.
Sc2ts is implemented in Python,
building on core PyData infrastructure~\cite{Harris2020array,Virtanen2020scipy,
Mckinney2010data,Lam2015numba} and bioinformatics 
packages\cite{Shirley2015efficient,Kunzmann2018biotite,Kunzmann2023biotite}.
See STAR Methods for details on all aspects of the method.

@@

Furthermore, some of the Pango X lineages above originate from
shared recombination events in the ARG (Table~\ref{tab:pango_x_lineages}).
An example of a set of ``nested'' closely related lineages
consists of XZ, XAC, XAD, XAE, and XAP,
which descend from a single recombination event in the ARG, rather than five independent
ones (Figure~\ref{fig:subgraphs}B), and
which we tentatively group using the unused label Xx.
These Pango X lineages were designated around the same time.
These results highlight that the ARG can help
clarify the relationships between recombinants,
allowing more precise interpretation of evidence for recombination.
